---
title: "Practical machine learning - Course project"
author: "Celine B"
date: "08/10/2020"
output: html_document
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r libraries, results="hide"}
library (ggplot2)
library (caret)
library(rpart)
library(randomForest)
```

### 1. Synopsis
Background
Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

Data
The training data for this project are available here:https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv
The test data are available here:https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv
The data for this project come from this source: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har.

Goal
The goal of your project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

### 2. Data loading and pre-processing
First step is to download the datasets (if not already done) and then to read them

```{r download}
if(file.exists("training.csv")) {
    print("File already downloaded")
} else {
    trainingURL <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
    download.file(trainingURL, "training.csv")
}
if(file.exists("testing.csv")) {
    print("File already downloaded")
} else {
    testingURL <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
    download.file(testingURL, "testing.csv")
}

training <- read.csv("training.csv")
testing  <- read.csv("testing.csv")
```

Second step: cleaning. Variables with nearly zero variance are removed, idem for the ones that are mostly NA and for identification variables (columns 1:5)

```{r clean}
nzv <- nearZeroVar(training)
training <- training[, -nzv]
testing <- testing[, -nzv]

mostlyNA <- sapply(training, function(x) mean(is.na(x))) > 0.95
training <- training[, mostlyNA==F]
testing <- testing[, mostlyNA==F]

training <- training[, -(1:5)]
testing  <- testing[, -(1:5)]
```

```{r partition}
inTrain  <- createDataPartition(training$classe, p=0.7, list=FALSE)
TrainSet <- training[inTrain, ]
ValidSet <- training[-inTrain, ]
```

### 3. Evaluation of several prediction models

#### Random forest model
The Random Forest method is applied on the training data with the 'classe' variable as outcome and the prediction made based on other variables in the data.

```{r randomForest}
modFit <- randomForest(classe ~ ., data=TrainSet)
modFit
```

Prediction using Random forest
```{r predictrandomForest}
predict <- predict(modFit, ValidSet, type="class")
confusionMatrix(ValidSet$classe, predict)
```

#### Decision tree
Using a second model, the decision tree.

```{r decisionTree}
modFit2 <- rpart(classe~., TrainSet)
```

Prediction using Decision tree
```{r predictdecisionTree}
predict2 <- predict(modFit2, ValidSet, type="class")
confusionMatrix(ValidSet$classe, predict2)
```

#### Generalized Boosted Model (GBM)
Third model to be assessed: the generalised boosted model.

```{r GBM}
control <- trainControl(method = "repeatedcv", number=5, repeats=1)
modFit3 <- train(classe~., TrainSet, method="gbm", trControl=control, verbose=FALSE)
```

Prediction using GBM

```{r predictGBM}
predict3 <- predict(modFit3, ValidSet)
confusionMatrix(predict3, ValidSet$classe)
```

### 4. Model selection
In the three above models, we will select the one that has the highest accuracy value and consequently the lowest error rate. From the results above, this corresponds to the random forest model.
We will now apply that model to the initial testing set.

```{r predicttesting}
predictF <- predict(modFit, testing, type="class")
predictF
```